---
title: "SBTN State of Nature Water unified layers"
subtitle: "water availability and water pollution"
author: "Rafael Camargo"
date: "July 2, 2024"
engine: knitr
format:
  html:
    toc: true
execute:
  warning: false    
---

## 1. Setup

Load required packages

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here, dplyr, magrittr, sf, terra, stars, exactextractr, skimr, readr, tidyr, openxlsx, mapview)
```

```{r}
sf::sf_use_s2(FALSE) # due to issues in HydroBASINS geometries
```

## 2. Load data

### 2.1. Water Availability

#### 2.1.1. Baseline Water Stress [Kuzma et al. 2023](https://www.wri.org/data/aqueduct-global-maps-40-data)

```{r}
bws <- read_sf(here("inputs", "aqueduct40", "GDB", "Aq40_Y2023D07M05.gdb"), layer = "baseline_annual")
```

#### 2.1.2. Water Depletion [Brauman et al. 2016](http://www.earthstat.org/water-depletion-watergap3-basins/)

```{r}
wdp <- rast(here("inputs", "water_depletion", "WaterDepletionCat_WG3.tif"))
```

#### 2.1.3. Blue Water Reduction [Hogeboom et al. 2024](https://data.4tu.nl/datasets/089afa18-bbd5-493e-8a6c-39e1ca7b2a59)

```{r}
bwr <- read_sf(here("inputs", "sbtn_freshwater_quantity_target_h5", "Freshwater Quantity Targets.shp"))
```

### 2.2. Water Pollution

#### 2.2.1. Coastal Eutrophication Potential [Kuzma et al. 2023](https://www.wri.org/data/aqueduct-global-maps-40-data)

```{r}
cep <- bws
```

#### 2.2.2. Nitrate-Nitrite Concentration [Damania et al. 2019](https://wbwaterdata.org/dataset/global-nitrate-nitrite-surface-water)

```{r}
nox_csv <- read_csv(here("inputs", "nitrate_nitrite", "noxn_preds_05_16_19.csv"))
```

*Spatial grid*
```{r}
sp_grid <- read_csv(here("inputs", "nitrate_nitrite", "ML_data_gridcell_match.csv"))
```

#### 2.2.3. Periphyton Growth Potential [McDowell et al. 2020](https://figshare.com/s/e44fe7319585183d888f)

*Total Nitrogen Concentration*
```{r}
tnc <- rast(here("inputs", "periphyton_growth_potential", "TN_Conc.tif"))
```

*Total Phosphorus Concentration*
```{r}
tpc <- rast(here("inputs", "periphyton_growth_potential", "TP_Conc.tif"))
```

### 2.3. Aggregation unit

#### 2.3.1. WMO Basins and Sub-Basins [GRDC 2020](https://panda.maps.arcgis.com/home/item.html?id=be4b6f13121b4670ad8f006bc1908e14)

*To add name of basins*
```{r}
# wmo_basins <- st_read(here("inputs", "wmo_basins", "wmobb_basins.shp")) |> 
#   select(WMOBB_id = WMOBB, WMOBB_name = WMOBBName, region = REGNAME)
```

#### 2.3.2. HydroSHEDS HydroBASINS Level 6 [Lehner & Grill 2013](https://www.hydrosheds.org/products/hydrobasins)

```{r}
# h6 <- st_read(here("inputs", "hydrosheds", "hybas_lev06_v1c.shp")) |>
#   select(HYBAS_ID) |>
#   mutate(HYBAS_ID = as.factor(HYBAS_ID)) |>
#   st_join(wmo_basins, largest = TRUE)
```

#### 2.3.3. Export / Fast load

```{r}
# write_rds(h6, here("outputs", "h6.rds"))
h6 <- read_rds(here("outputs", "h6.rds"))
```

## 3. Clean and Classify data to 1-5 values

### 3.1. Water Availability

#### 3.1.1. Baseline Water Stress

```{r}
bws_n <- bws |>
  mutate(
    bws_raw = na_if(bws_raw, -9999),
    bws_raw = na_if(bws_raw, 9999),
    bws_n = case_when(
      bws_label == "Low (<10%)" ~ 1,
      bws_label == "Low - Medium (10-20%)" ~ 2,
      bws_label == "Medium - High (20-40%)" ~ 3,
      bws_label == "High (40-80%)" ~ 4,
      bws_label %in% c("Extremely High (>80%)", "Arid and Low Water Use") ~ 5
    )
  ) |>
  select(bws_raw, bws_n, bws_label, geometry = Shape)
```

#### 3.1.2. Water Depletion

```{r}
wdp_reclass <- matrix(
  c(
    0,NA,
    1,1, # <=5% annual depletion
    2,2, # >5–25% annual depletion
    5,3, # >25–75% annual depletion, Dry-Year
    6,4, # >25–75% annual depletion, Seasonal
    7,5, # >75–100% annual depletion
    8,5  # >100% annual depletion
  ),
  ncol=2, byrow=TRUE
)
```

```{r}
wdp_n <- wdp |>
  classify(wdp_reclass) |> 
  exact_extract(h6, fun = c("mean", "median", "quantile", "majority", "coefficient_of_variation"), quantiles = 0.75,
                       append_cols = "HYBAS_ID", progress = TRUE) 

skim(wdp_n)
```

```{r}
wdp_n <- wdp_n |>
  mutate(
    wdp_raw = median,
    wdp_n = round(median),
    wdp_label = case_when(
      wdp_n == 1 ~ "<=5% annual depletion",
      wdp_n == 2 ~ ">5–25% annual depletion",
      wdp_n == 3 ~ ">25–75% annual depletion, Dry-Year",
      wdp_n == 4 ~ ">25–75% annual depletion, Seasonal",
      wdp_n == 5 ~ ">75% annual depletion"
    )
  ) |>
  select(HYBAS_ID, wdp_raw:wdp_label)
```

#### 3.1.3. Blue Water Reduction

```{r}
bwr_n <- bwr |>
  mutate(
    bwr_raw = FQT_Annual,
    bwr_n = case_when(
      bwr_raw < 5 ~ 1,
      bwr_raw < 25 ~ 2,
      bwr_raw < 50 ~ 3,
      bwr_raw < 80 ~ 4,
      bwr_raw >= 80 ~ 5
    ),
    bwr_label = case_when(
      bwr_n == 1 ~ "Low (<5%)",
      bwr_n == 2 ~ "Moderate (5-25%)",
      bwr_n == 3 ~ "Significant (25-50%)",
      bwr_n == 4 ~ "Severe (50-80%)",
      bwr_n == 5 ~ "Extreme (>=80%)"
    )
  ) |>
  select(bwr_raw, bwr_n, bwr_label)
```

#### 3.1.4. Join datasets and Export

```{r}
wa_n <- h6 |>
  st_join(bws_n, largest = TRUE) |>
  left_join(wdp_n, by = "HYBAS_ID") |>
  st_join(bwr_n, largest = TRUE)
```

```{r}
write_rds(wa_n, here("outputs", "wa_n.rds"))
```

### 3.2. Water Pollution

#### 3.2.1. Coastal Eutrophication Potential

```{r}
cep_n <- cep |>
  mutate(
    cep_raw = na_if(cep_raw, -9999),
    cep_raw = na_if(cep_raw, 9999),
    cep_n = case_when(
      cep_label == "Low (<-5)" ~ 1,
      cep_label == "Low - Medium (-5 to 0)" ~ 2,
      cep_label == "Medium - High (0 to 1)" ~ 3,
      cep_label == "High (1 to 5)" ~ 4,
      cep_label == "Extremely High (>5)" ~ 5
    )
  ) |>
  select(cep_raw, cep_n, cep_label, geometry = Shape)
```

#### 3.2.2. Nitrate-Nitrite Concentration

*Get mean predicted values between 2006 and 2010, then rasterize*
```{r}
nox <- nox_csv |>
  filter(year >= 2006) |>
  group_by(OBJECTID) |>
  summarise(value = mean(preds)) |>
  right_join(sp_grid, by = "OBJECTID") |>
  select(-OBJECTID) |>
  st_as_sf(coords = c("Lon", "Lat"), crs = 4326) |>
  st_rasterize(dx = 0.5, dy = 0.5) |>
  rast()
```

```{r}
nox_n <- exact_extract(nox, h6, fun = c("mean", "median", "quantile", "majority", "coefficient_of_variation"), quantiles = 0.75,
                       append_cols = "HYBAS_ID", progress = TRUE)

skim(nox_n)
```

```{r}
nox_n <- nox_n |>
  mutate(
    nox_raw = median,
    nox_n = case_when(
      median <= 0.4 ~ 1,
      between(median, 0.4, 0.8) ~ 2,
      between(median, 0.8, 1.2) ~ 3,
      between(median, 1.2, 1.6) ~ 4,
      median > 1.6 ~ 5
    ),
    nox_label = case_when(
      nox_n == 1 ~ "Very Low Concentration (<=0.4)",
      nox_n == 2 ~ "Low Concentration (>0.4-0.8)",
      nox_n == 3 ~ "Moderate Concentration (>0.8-1.2)",
      nox_n == 4 ~ "High Concentration (>1.2-1.6)",
      nox_n == 5 ~ "Very High Concentration (>1.6)"
    )
  ) |>
  select(HYBAS_ID, nox_raw:nox_label)
```

#### 3.2.3. Periphyton Growth Potential

```{r}
tnc_n <- exact_extract(tnc, h6, fun = c("mean", "median", "quantile", "majority", "coefficient_of_variation"), quantiles = 0.75,
                       append_cols = "HYBAS_ID", progress = TRUE)

skim(tnc_n)
```

```{r}
tpc_n <- exact_extract(tpc, h6, fun = c("mean", "median", "quantile", "majority", "coefficient_of_variation"), quantiles = 0.75,
                       append_cols = "HYBAS_ID", progress = TRUE)

skim(tpc_n)
```

```{r}
pgp_n <- tnc_n |>
  select(HYBAS_ID, tnc_raw = median) |>
  left_join(
    select(tpc_n, HYBAS_ID, tpc_raw = median),
    by = "HYBAS_ID" 
  ) |>
  mutate(
    ratio_n_p = tnc_raw/tpc_raw,
    limiting = if_else(ratio_n_p < 7, "N-limited", "P-limited"),
    pgp_n = case_when(
      limiting == "N-limited" & tnc_raw <= 0.4 ~ 1,
      limiting == "N-limited" & between(tnc_raw, 0.4, 0.8) ~ 2,
      limiting == "N-limited" & between(tnc_raw, 0.8, 1.2) ~ 3,
      limiting == "N-limited" & between(tnc_raw, 1.2, 1.6) ~ 4,
      limiting == "N-limited" & tnc_raw > 1.6 ~ 5,
      
      limiting == "P-limited" & tpc_raw <= 0.023 ~ 1,
      limiting == "P-limited" & between(tpc_raw, 0.023, 0.046) ~ 2,
      limiting == "P-limited" & between(tpc_raw, 0.046, 0.100) ~ 3,
      limiting == "P-limited" & between(tpc_raw, 0.100, 0.150) ~ 4,
      limiting == "P-limited" & tpc_raw > 0.150 ~ 5
    ),
    pgp_label = case_when(
      limiting == "N-limited" & pgp_n == 1 ~ "N-limited growth acceptable",
      limiting == "N-limited" & pgp_n == 2 ~ "N-limited growth acceptable",
      limiting == "N-limited" & pgp_n == 3 ~ "N-limited growth undesirable",
      limiting == "N-limited" & pgp_n == 4 ~ "N-limited growth undesirable",
      limiting == "N-limited" & pgp_n == 5 ~ "N-limited growth undesirable",
      
      limiting == "P-limited" & pgp_n == 1 ~ "P-limited growth acceptable",
      limiting == "P-limited" & pgp_n == 2 ~ "P-limited growth acceptable",
      limiting == "P-limited" & pgp_n == 3 ~ "P-limited growth undesirable",
      limiting == "P-limited" & pgp_n == 4 ~ "P-limited growth undesirable",
      limiting == "P-limited" & pgp_n == 5 ~ "P-limited growth undesirable"
    )
  ) |>
  select(HYBAS_ID, tnc_raw, tpc_raw, ratio_n_p, limiting, pgp_n, pgp_label)
```

#### 3.2.4. Join datasets and Export

```{r}
wp_n <- h6 |>
  st_join(cep_n, largest = TRUE) |>
  left_join(nox_n, by = "HYBAS_ID") |>
  left_join(pgp_n, by = "HYBAS_ID")
```

```{r}
write_rds(wp_n, here("outputs", "wp_n.rds"))
```

## 4. Identify the value for the unified layer

### 4.1. Max of Water Availability

```{r}
wa_n <- read_rds(here("outputs", "wa_n.rds"))
```

```{r}
son_wa <- wa_n |>
  rowwise() |>
  mutate(
    wa_max = max(c(bws_n, wdp_n, bwr_n), na.rm = TRUE),
    wa_max = if_else(between(wa_max, 1, 5), wa_max, NA),
    .after = region
  ) |>
  ungroup()
```

### 4.2. Max of Water Pollution

```{r}
wp_n <- read_rds(here("outputs", "wp_n.rds"))
```

```{r}
son_wp <- wp_n |>
  rowwise() |>
  mutate(
    wp_max = max(c(cep_n, nox_n, pgp_n), na.rm = TRUE),
    wp_max = if_else(between(wp_max, 1, 5), wp_max, NA),
    .after = region
  ) |>
  ungroup()
```

## 5. Join final output

```{r}
son_water <- son_wa |>
  left_join(
    son_wp |> select(-c(WMOBB_id:region)) |> st_drop_geometry(),
    by = "HYBAS_ID"
  ) |>
  relocate(geometry, .after = last_col())

st_drop_geometry(son_water) |> skim()
```

```{r}
son_water |> 
  st_transform("EPSG:8857") |> 
  st_simplify(dTolerance = 10000) |> 
  st_transform(st_crs(son_water)) |> 
  mutate(across(c(wa_max, wp_max), ~ as.factor(.x))) |> 
  mapview(
    layer.name = list("Max Water Availability", "Max Water Pollution"),
    zcol = c("wa_max", "wp_max"),
    col.regions = c("#FFFF99", "#FFE600", "#FF9900", "#FF1900", "#990000"),
    lwd = 0.2,
    alpha.regions = 0.9
  )
```

## 6. Export

```{r}
write_rds(son_water, here("outputs", "sbtn_son_water_v2.rds"))
```

```{r}
st_write(son_water, here("outputs", "sbtn_son_water_v2.shp"), layer_options = "ENCODING=UTF-8", delete_layer = TRUE)
```

```{r}
write.xlsx(list(
  "SBTN SoN Water" = st_drop_geometry(son_water)
), file = here("outputs", "sbtn_son_water_v2.xlsx"), rowNames = FALSE)
```
